Exploring Zstandard user-provided dictionary compression for FASTA files

### 2022-12-05 paper/paper.qmd
Start Methods, Discussion sections

  Zstandard (Zstd) represents a universal, lossless data compression standard and implementation that is highly configurable and is aimed at coupling high compression ratios with fast compression/decompression performance. Previous studies have paired specific Zstd configurations with various file formats in bioinformatics to reduce total data volume. This paper presents a "training mode" pipeline, written in the Julia programming language, wherein a custom compression dictionary is generated from a sample FASTA set in order to explore further compression improvements and compare them to the compression performance of Xz, Zlib, Bzip2, and Lz4 universal compressors.

@@

A list of all direct and indirect dependencies can be found in the repository's Manifest.toml file.

@@

# Conclusion

>>>>

  Zstandard (Zstd) represents a universal, lossless data compression standard and implementation that is highly configurable and is aimed at coupling high compression ratios with fast compression/decompression performance. Previous studies have paired specific Zstd configurations with various file formats in bioinformatics to reduce total data volume. This paper presents a primitive "training mode" model, written in the Julia programming language, wherein a custom compression dictionary is generated from a sample FASTA set in order to explore further compression improvements and compare them to the compression performance of Xz, Zlib, Bzip2, and Lz4 universal compressors.

@@

For the model implementation, Julia was chosen as there is package support for working with sequence data as well as for trivial transcoding of compressed or decompressed data via the TranscodingStreams package wherein compression algorithm implementations are loaded as codec I/O streams[@sato]. A list of all direct dependencies can be found in the repository's Project.toml file. The Julia implementation operates by first unloading a tarball containing the set of FASTA training files into an artifact, or life-cycled datastore.

@@

The provided FASTA training dataset was retrieved from a set of \textit{E.coli} MS 200-1 strain genomic scaffolds from a whole genome shotgun run[@ncbi]. Sampling a single organism may allow for additional compression/decompression improvements as the sequence data could possess repeating patterns unique to that individual's or species' genome that are recognized and added to the Zstd dictionary during training. The FASTA files themselves may bear similarity with one another in the same set, such as each file being of similar length. Simultaneously, the Zstd dictionary may not achieve comparable compression/decompression performance when attempted on more dissimilar sequence data. Portability of Zstd dictionaries, in the sense of being universally applicable across all sequence data, could be weighed against generating dedicated dictionaries for specific datasets depending on the storage requirements.  Zstd parameters were kept default, though they can be manually assigned, including advanced compression options like compression job size or selecting from 22 predefined compression levels[@facebook].

### 2022-12-05 paper/paper.tex
Start Methods, Discussion sections

Zstandard (Zstd) represents a universal, lossless data compression
standard and implementation that is highly configurable and is aimed at
coupling high compression ratios with fast compression/decompression
performance. Previous studies have paired specific Zstd configurations
with various file formats in bioinformatics to reduce total data volume.
This paper presents a ``training mode'' pipeline, written in the Julia
programming language, wherein a custom compression dictionary is
generated from a sample FASTA set in order to explore further
compression improvements and compare them to the compression performance
of Xz, Zlib, Bzip2, and Lz4 universal compressors.

@@

A list of all direct and indirect dependencies can be found in the
repository's Manifest.toml file.

@@

\hypertarget{conclusion}{%
\section{Conclusion}\label{conclusion}}

>>>>

Zstandard (Zstd) represents a universal, lossless data compression
standard and implementation that is highly configurable and is aimed at
coupling high compression ratios with fast compression/decompression
performance. Previous studies have paired specific Zstd configurations
with various file formats in bioinformatics to reduce total data volume.
This paper presents a primitive ``training mode'' model, written in the
Julia programming language, wherein a custom compression dictionary is
generated from a sample FASTA set in order to explore further
compression improvements and compare them to the compression performance
of Xz, Zlib, Bzip2, and Lz4 universal compressors.

@@

For the model implementation, Julia was chosen as there is package
support for working with sequence data as well as for trivial
transcoding of compressed or decompressed data via the
TranscodingStreams package wherein compression algorithm implementations
are loaded as codec I/O streams\citep{sato}. A list of all direct
dependencies can be found in the repository's Project.toml file. The
Julia implementation operates by first unloading a tarball containing
the set of FASTA training files into an artifact, or life-cycled
datastore.

@@

The provided FASTA training dataset was retrieved from a set of
\textit{E.coli} MS 200-1 strain genomic scaffolds from a whole genome
shotgun run\citep{ncbi}. Sampling a single organism may allow for
additional compression/decompression improvements as the sequence data
could possess repeating patterns unique to that individual's or species'
genome that are recognized and added to the Zstd dictionary during
training. The FASTA files themselves may bear similarity with one
another in the same set, such as each file being of similar length.
Simultaneously, the Zstd dictionary may not achieve comparable
compression/decompression performance when attempted on more dissimilar
sequence data. Portability of Zstd dictionaries, in the sense of being
universally applicable across all sequence data, could be weighed
against generating dedicated dictionaries for specific datasets
depending on the storage requirements. Zstd parameters were kept
default, though they can be manually assigned, including advanced
compression options like compression job size or selecting from 22
predefined compression levels\citep{facebook}.
